---
title: "Homework 3"
date: "Due February 2, 2017 at 5pm in Snedecor 2404"
output: 
  pdf_document:
    number_sections: true
geometry: margin=1in
---

Please show all work for full credit. Print and staple your assignment together and submit by 5pm of the due date in Snedecor 2404. If you cannot attend class or office hours on the due date, please arrange to submit your homework prior to the due date.

```{r setup, echo=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(xtable)

knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2)
theme_set(theme_bw(base_family = "serif"))
```

1.  [Ch. 3.1 Exercise 3, pg. 77] Osborne, Bishop, and Klein collected manufacturing data on the torques required to loosed bolts holding an assembly on a piece of heavy machinery. The accompanying table (also available on the website as `bolts.csv`) shows part of their data concerning two particular bolts. The torques recorded (in ft lb) were taken from 15 different pieces of equipment as they were assembled.
    a) Make a scatterplot of these paired data. Are there any obvious patterns in the data?
    b) A trick often employed in the analysis of pared data such as these is to reduce the pairs to differences by subtracting the values of one of the variables from the other. Compute differences (top bolt - bottom bolt) here. Then make and interpret a dot diagram for these values.
            
        ```{r bolts}
        bolts <- read_csv("../data/bolts.csv")
        
        kable(bolts)
        ```
        
2. [Ch 3, Exercise 3, pg. 114] The accompanying data (also available on the website as `manganese.csv`) are three hypothetical samples of size 10 that are supposed to represent measured manganese contents in specimens of 1045 steel (the units are points, or .01%). Suppose that these measurements were made on standard specimens having "true" manganese contents of 80, using three different analytical methods. (Thirty specimens were involved.)

    
    ```{r manganese}
    manganese <- read_csv("../data/manganese.csv")
    
    kable(t(manganese), col.names = rep("", 10))
    ```
    
    a) Make (on the same coordinate system) side by side boxplots that you can use to compare the three analytical methods.
    b) Discuss the apparent effectiveness of the three methods in terms of the appearance of your diagram from a) and in terms of the concepts of accuracy and precision discussed in Section 1.3 of the notes.
    c) An alternative method of comparing two such analytical methods is to use both methods of analysis once on each of (say) 10 different specimens (10 specimens and 20 measurements). The the terminology of Section 1.2, what kind of data would be generated by such a plan? If one simply wishes to compare the average measurements produced by two analytical methods, which data collection plan (20 specimens and 20 measurements, or 10 specimens and 20 measurements) seems to you most likely to provide the better comparison? Explain.
    
3. [Ch 3, Exercise 8, pg. 116] The accompanying data are the times to failure (in millions of cycles) of high-speed turbine engine bearings made out of two different compounds. These were taking from "Analysis of Single Classification Experiments Based on Censored Samples from the Two-parameter Weibull Distribution" by J. I. McCool (*The Journal of Statistical Planning and Inference*, 1979).
    
    ```{r failure}
    failure <- data.frame(compound_1 = c(3.03, 5.53, 5.60, 9.30, 9.92, 12.51, 12.95, 15.21, 16.04, 16.84),
                          compound_2 = c(3.19, 4.26, 4.47, 4.53, 4.67, 4.69, 5.78, 6.79, 9.37, 12.75))
    
    kable(t(failure), col.names = rep("", nrow(failure)))
    ```
    
    a) Find the .84 quantile of the Compound 1 failure times.
    b) Give the coordinates of the two lower-left points that would appear on a normal plot of the Compound 1 data.
    c) Make back-to-back stem-and-leaf plots for comparing the life length properties of bearings made from Compounds 1 and 2.
    d) Make (to scale) side-by-side boxplots for comparing the life length for the two compounds. Mark numbers on the plots indicating the locations of their main features.
    e) Compute the sample means and standard deviations of the two sets of lifetimes.
    f) Describe what your answers to parts c), d), and e) above indicate about the life lengths of these turbine bearings.
    
4. [Ch 3, Exercise 17, pg. 119] The data in the accompanying table are measurements of the latent heat of fusion of ice taken from *Experimental Statistics* (NBS Handbook 91) by M. G. Natrella. The measurements were made (on specimens cooled to $-.072$°C) using two different methods. The first was an electrical method, and the second was a method of mixtures. The units are calories per gram of mass.

    ```{r heat}
    heat <- data.frame(electrical = c(79.98, 80.04, 80.02, 80.04, 80.03, 80.03, 80.04, 79.97, 80.05, 80.03, 80.02, 80.00, 80.02),
                       mixtures = c(80.02, 79.94, 79.98, 79.97, 79.97, 80.03, 79.95, 79.97, rep("", 5)))
    
    kable(t(heat), col.names = rep("", nrow(heat)))
    ```
    
    a) Make side-by-side boxplots for comparing the two measurement methods. Does there appear to be any important difference in the precision of the two methods? Is it fair to say that at least one of the methods must be somewhat inaccurate? Explain.
    
    b) Compute and compare the sample means and the sample standard deviations for the two methods. How are the comparisons of these numerical quantities already evident on your plot in a)?
    


