---
output: 
  pdf_document:
    keep_tex: true
    number_sections: true
    includes:
      in_header: 00_header.tex
fontsize: 12pt
geometry: margin=1in
linestretch: 1.5
---

```{r setup, echo=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(xtable)
library(ggplot2)

knitr::opts_chunk$set(echo=FALSE, message=FALSE, fig.height = 3)
theme_set(theme_bw(base_family = "serif"))
```

\setcounter{section}{0}

# Introduction

First we start with a roadmap for the course. The subject of **engineering statistics** is defined, its importance is described and basic terminology is introduced.

## What is statistics?

Statistics is the science of collecting, presenting, analyzing, and making decisions from data. Often, as an engineer, it is necessary to **collect and interpret data** that will help in understanding how a new system or product works. 

Statistics has applications to engineering through quality control, process control, reliability, risk management, system identification, design of experiments, etc.

\vspace{.25in}
\begin{df}
\emph{Engineering statistics} is the study of how best to 
\begin{enumerate}
\item collect engineering data,
\item summarize or describe engineering data, and
\item draw formal inference and practical conclusions on the basis of engineering data,
\end{enumerate}
all while recognizing the reality of variation.
\end{df}
\vspace{.1in}

We can break down this study into three main tasks:

\begin{enumerate}
\item {\bf Summary:} 
\item {\bf Inference:}
\item {\bf Interpretation:}
\end{enumerate}

\newpage

```{r gears-load}
gears <- read_csv("../../data/gears.csv")
```



\begin{ex}[Heat treating gears, pg. 2]
A process engineer is faced with the question, "How should gears be loaded into a continuous carburizing furnace in order to minimize distortion during heat treating?" The engineer conducts a well-thought-out study and obtains the runout values for `r gears %>% filter(Arrangement == "laid") %>% nrow()` gears laid and `r gears %>% filter(Arrangement == "hung") %>% nrow()` gears hung.

```{r, results='asis'}
gears %>%
  group_by(Arrangement) %>%
  summarise(Values = paste(Runout, collapse = ", ")) %>%
  spread(Arrangement, Values) %>%
  xtable(align = c("c", "p{1in}", "p{1in}"), caption = "Thrust face runouts (.0001 in.)") %>%
  print(include.rownames = FALSE)
```

This data, as is, is hard to get insights from. Should the gears be hung or laid? We still cannot tell by looking at the table.

{\bf Numerical summaries:} 
\vspace{.75in}

{\bf Variation:} 
```{r, fig.cap="Dot diagrams of runouts"}
gears %>%
  ggplot() +
  geom_dotplot(aes(Runout)) +
  facet_wrap(~Arrangement)
```

\end{ex}




## Populations and samples

## Experimental vs. observational studies

## Data and measurement

## Complete/Fractional factorial designs
