---
output: 
  pdf_document:
    number_sections: true
    includes:
      in_header: 00_header.tex
fontsize: 12pt
fig_caption: true
geometry: margin=1in
linestretch: 1.5
---

```{r setup, echo=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(xtable)
library(MASS)

knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2)
theme_set(theme_bw(base_family = "serif"))

set.seed(305)
```

\setcounter{section}{5}
\setcounter{subsection}{1}
\setcounter{ex}{16}
\setcounter{df}{11}

## Continuous random variables

It is often convenient to think of a random variable as having a whole (continuous) interval for its set of possible values.

The devices used to describe continuous probability distributions differ from those that describe discrete probability distributions.

Examples of continuous random variables:

\newpage

### Probability density functions and cumulative distribution functions

A *probability density function (pdf)* is the continuous analogue of a discrete random variable's probability mass function (pmf).

\vspace{.1in}
\begin{df}
A \emph{probability density function (pdf)} for a continuous random variable $X$ is a nonnegative function $f(x)$ with
$$
\int\limits_{-\infty}^{\infty} f(x) = 1
$$
and such that for all $a \le b$,
$$
P[a \le X \le b] = \int\limits_a^bf(x)dx.
$$
\end{df}

\vspace{.2in}

1. 
2. 
3. 

\vfill

```{r, fig.height=3}
x <- seq(-5, 22, length.out = 200)
f <- .6*dnorm(x, 3, 2) + .3*dnorm(x, 9, 1) + .1*dnorm(x, 19, .4)
shade <- rbind(data.frame(x = x[x >= 2 & x <= 6], y = f[x >= 2 & x <= 6]),
               data.frame(x = rev(x[x >= 2 & x <= 6]), y = 0))

qplot(x, f, geom = "line") +
  geom_polygon(aes(x, y), data = shade, fill = "blue", alpha = .4) +
  ylab(expression("f(x)"))
```

\newpage

\begin{ex}[Compass needle]
Consider a de-magnetised compass needle mounted at its centre so that it can spin freely.  It is spun clockwise and when it comes to rest the angle, $\theta$, from the vertical, is measured. Let 
$$
Y = \text{the angle measured after each spin in radians}
$$

\vspace{.2in}

What values can $Y$ take?

\vspace{.2in}

What form makes sense for $f(y)$?

\vspace{3in}

If this form is adopted, that what must the pdf be?

\newpage

Using this pdf, calculate the following probabilities:

\begin{enumerate}
\itemsep 1.5in 
\item $P[Y < \frac{\pi}{2}]$
\item $P[\frac{\pi}{2} < Y < 2\pi]$
\item $P[\frac{\pi}{6} < Y < \frac{\pi}{4}]$
\item $P[Y = \frac{\pi}{6}]$
\end{enumerate}
\vspace{1in}



\end{ex}

\newpage

\begin{df}
The \emph{cumulative distribution function (cdf)} of a continuous random variable $X$ is a function $F$ such that
$$
F(x) = P[X \le x] = \int\limits_{-\infty}^x f(t) dt
$$
\end{df}

$F(x)$ is obtained from $f(x)$ by integrtion, and applying the fundamental theorem of calculus yields

$$
\frac{d}{dx}F(x) = f(x).
$$

That is, $f(x)$ is obtained from $F(x)$ by differentiation.

As with discrete random variables, $F$ has the following properties:

\begin{enumerate}
\itemsep 1in
\item
\item
\item
\end{enumerate}

\newpage

\begin{ex}[Compass needle, cont'd]
Recall the compass needle example, with
$$
f(x) = \begin{cases} \frac{1}{2\pi} & 0 \le y \le 2\pi \\ 0 & \text{otherwise}\end{cases}
$$
Find the cdf.

\vspace{.2in}
For $y < 0$

\vspace{1in}
For $0 \le y \le 2\pi$

\vspace{1in}
For $y > 2\pi$
\vfill


\newpage

Calculate the following using the cdf:

$F(1.5)$

\vfill

$P[Y \le \frac{4\pi}{5}]$

\vfill

$P[Y > 5]$

$P[\frac{\pi}{3} < Y \le \frac{\pi}{2}]$

\end{ex}

### Means and variances for continuous distributions

### Quantiles

### The Normal distribution

### Other special continuous distributions


