---
output: 
  pdf_document:
    number_sections: true
    includes:
      in_header: 00_header.tex
fontsize: 12pt
fig_caption: true
geometry: margin=1in
linestretch: 1.5
---

```{r setup, echo=FALSE, message=FALSE}
library(knitr)
library(tidyverse)
library(xtable)
library(MASS)

knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2)
theme_set(theme_bw(base_family = "serif"))

set.seed(305)
```

\setcounter{section}{6}
\setcounter{subsection}{2}
\setcounter{ex}{10}
\setcounter{df}{2}
\setcounter{page}{20}

## Hypothesis testing

Last section illustrated how probability can enable confidence interval estimation. We can also use probability as a means to use data to quantitatively assess the plausibility of a trial value of a parameter.

\vspace{.2in}

**Statistical inference** is using data from the sample to draw conclusions about the population.

1. Interval estimation (confidence intervals)

\vspace{.5in}

2. Hypothesis testing

\vspace{.5in}

\begin{df}
Statistical \emph{significance testing} is the use of data in th quantitative assessment of the plausibility of some trial value for a parameter (or function of one or more parameters).
\end{df}

\vspace{.2in}

Significance (or hypothesis) testing begins with the specification of a trial value (or **hypothesis**).

\vspace{.2in}

\begin{df}
A \emph{null hypothesis} is a statement of the form
$$
\text{Parameter}=\#
$$
or
$$
\text{Function of parameters}=\#
$$
for some $\#$ that forms the basis of investigation in a significance test. A null hypothesis is usually formed to embody a status quo/"pre-data" view of the parameter. It is denoted $\text{H}_0$.
\end{df}
\newpage

\begin{df}
An \emph{alternative hypothesis} is a statement that stands in opposition to the null hypothesis. It specifies what forms of departure from the null hypothesis are of concern. An alternative hypothesis is denoted as $\text{H}_a$. It is of the form 
$$
\text{Parameter}\not=\# \quad \text{ or } \quad \text{Parameter}>\# \quad \text{ or } \quad \text{Parameter}<\# \quad 
$$
\end{df}

\vspace{.1in}

Examples (testing the true mean value):
\begin{eqnarray*}
\text{H}_0: \mu = \# & \text{H}_0: \mu = \# & \text{H}_0: \mu = \# \\
\text{H}_a: \mu \not= \# & \text{H}_a: \mu > \# & \text{H}_a: \mu < \#
\end{eqnarray*}

Often, the alternative hypothesis is based on an investigator's suspicions and/or hopes about th true state of affairs.

\vspace{.2in}

The **goal** is to use the data to debunk the null hypothesis in favor of the alternative.

1. Assume $\text{H}_0$.
2. Try to show that, under $\text{H}_0$, the data are preposterous.
3. If the data are preposterous, reject $\text{H}_0$ and conclude $\text{H}_a$.

The outcomes of a hypothesis test consists of:

\newpage

\begin{ex}[Fair coin]
Suppose we toss a coin $n=25$ times, and the results are denoted by $X_1,X_2,\dots,X_{25}$. We use $1$ to denote the result of a head and $0$ to denote the results of a tail. Then $X_1 \sim Binomial(1,\rho)$ where $\rho$ denotes the chance of getting heads, so $\text{E}(X_1) = \rho, \text{Var}(X_1) = \rho(1-\rho)$. Given  the result is you got all heads, do you think the coin is fair?
\end{ex}

\vfill

In the real life, we may have data from many different kinds of distributions! Thus we need a universal framework to deal with these kinds of problems.

\vfill

\newpage

### Significance tests for a mean

\begin{df}
A \emph{test statistic} is the particular form of numerical data summarization used in a significance test.
\end{df}

\vspace{.2in}

\begin{df}
A \emph{reference (or null) distribution} for a test statistic is the probability distribution describing the test statistic, provided the null hypothesis is in fact true.
\end{df}

\vspace{.2in}

\begin{df}
The \emph{observed level of significance or $p$-value} in a significance test is the probability that the reference distribution assigns to the set of possible values of the test statistic that are at least as extreme as the one actually observed.
\end{df}

\newpage


Based on our results from Section 6.2 of the notes, we can develop hypothesis tests for the true mean value of a distribution in various situations, given an iid sample $X_1, \dots, X_n$ where $\text{H}_0: \mu = \mu_0$.

Let $K$ be the value of the test statistic, $Z\sim N(0,1)$, and $T\sim t_{n - 1}$. Here is a table of $p$-values that you should use for each set of conditions and choice of $\text{H}_a$.

\begin{table}[H]
\centering
\begin{tabular}{l | c | c c c} 
Situation & K & $\text{H}_a:\mu \not=\mu_0$ & $\text{H}_a:\mu < \mu_0$ & $\text{H}_a:\mu > \mu_0$ \\
\hline
$n \ge 25, \sigma \text{ known}$ & $\frac{\overline{x} - \mu_0}{\sigma/\sqrt{n}}$ & $P(|Z| > K)$ & $P(Z < K)$ & $P(Z > K)$ \\
$n \ge 25, \sigma \text{ unknown}$ & $\frac{\overline{x} - \mu_0}{s/\sqrt{n}}$ & $P(|Z| > K)$ & $P(Z < K)$ & $P(Z > K)$ \\
$n < 25, \sigma \text{ unknown}$ & $\frac{\overline{x} - \mu_0}{s/\sqrt{n}}$ & $P(|T| > K)$ & $P(T < K)$ & $P(T > K)$
\end{tabular}
\end{table}

Steps to perform a hypothesis test:

1.
\vfill
2.
\vfill
3.
\vfill
4.
\vfill
5.
\vfill
6.

\vfill
\newpage

\begin{ex}[Cylinders]
The strengths of $40$ steel cylinders were measured in MPa. The sample mean strength is $1.2$ MPa with a sample standard deviation of $0.5$ MPa. At significance level $\alpha = 0.01$, conduct a hypothesis test to determine if the cylinders meet the strength requirement of 0.8 MPa.
\end{ex}

\newpage

\begin{ex}[Concrete beams]
10 concrete beams were each measured for flexural strength (MPa). The data is as follows.

```{r, comment = NA, results='asis'}
beams <- c(8.2, 8.7, 7.8, 9.7, 7.4, 7.8, 7.7, 11.6, 11.3, 11.8)
print(beams)
```

The sample mean was $`r mean(beams)`$ MPa and the sample variance was $`r round(var(beams), 4)`$ MPa. Conduct a hypothesis test to find out if the flexural strength is different from $9.0$ MPa.



\end{ex}

\newpage


### Hypothesis testing using the CI

We can also use the $1-\alpha$ confidence interval to perform hypothesis tests (instead of $p$-values). The confidence interval will contain $\mu_0$ when there is little to no evidence against $\text{H}_0$ and will not contain $\mu_0$ when there is strong evidence against $\text{H}_0$.


Steps to perform a hypothesis test using a confidence interval:

1.
\vfill
2.
\vfill
3.
\vfill
4.
\vfill
5.
\vfill
6.

\vfill
\newpage

\begin{ex}[Breaking strength of wire, cont'd]
```{r}
wire <- c(100.37, 96.31, 72.57, 88.02, 105.89, 107.80, 75.84, 92.73, 67.47, 94.87, 122.04, 115.12, 95.24, 119.75, 114.83, 101.79, 80.90, 96.10, 118.51, 109.66, 88.07, 56.29, 86.50, 57.62, 74.7, 92.53, 86.25, 82.56, 97.96, 94.92, 62,93, 98.44, 119.37, 103.70, 72.40, 71.29, 107.24, 64.82, 93.51, 86.97)
```

Suppose you are a manufacturer of construction equipment. You make $0.0125$ inch wire rope and need to determine how much weight it can hold before breaking so that you can label it clearly. You have breaking strengths, in kg, for $`r length(wire)`$ sample wires with sample mean breaking strength $`r round(mean(wire), 2)`$ kg and sample standard deviation $`r round(sd(wire), 2)`$ kg. Using the appropriate $95$\% confidence interval, conduct a hypothesis test to find out if the true mean breaking strength is above $85$ kg.
\end{ex}


\newpage

\begin{ex}[Concrete beams, cont'd]
10 concrete beams were each measured for flexural strength (MPa). The data is as follows.

```{r, comment = NA, results='asis'}
beams <- c(8.2, 8.7, 7.8, 9.7, 7.4, 7.8, 7.7, 11.6, 11.3, 11.8)
print(beams)
```

The sample mean was $`r mean(beams)`$ MPa and the sample variance was $`r round(var(beams), 4)`$ MPa. At $\alpha= 0.01$, test the hypothesis that the true mean flexural strength is $10$ MPa using a confidence interval.

\end{ex}

\newpage

\begin{ex}[Paint thickness, cont'd]
Consider the following sample of observations on coating thickness for low-viscosity paint.

```{r, comment=NA, results='asis'}
paint <- c(0.83, 0.88, 0.88, 1.04, 1.09, 1.12, 1.29, 1.31, 1.48, 1.49, 1.59, 1.62, 1.65, 1.71, 1.76, 1.83)
paint
```

Using $\alpha= 0.1$, test the hypothesis that the true mean paint thickness is $1.00$ mm. Note, the $90$\% confidence interval for the true mean paint thickness was calculated from before as $(1.201,1.499)$.
\end{ex}

